---
title: "How to use 'StratoBayes'"
# author: 
#   - "Kilian Eichenseer <kilian.eichenseer@durham.ac.uk>"
#   - "Matthias Sinnesael <msinnesam@tcd.ie>"
#   - "Martin R. Smith <martin.smith@durham.ac.uk>"
#   - "Andrew Millard <a.r.millard@durham.ac.uk>"
# date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{StratoBayes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
  <!-- %\VignetteOption{rmarkdown.html_vignette.check_title=FALSE} -->
---

<script type="text/x-mathjax-config">
  MathJax.Hub.Config({
    "HTML-CSS": {
      preferredFont: "sans-serif",
      webFont: "sans-serif"
    }
  });
</script>

<style>
.mjx-chtml {
  font-family: sans-serif !important;
}
</style>
<!-- NOTES

- StratMap instead of AgeModel & AgeConversion (merge to 1 function for user)
- AgeConversion needs to check for vector input (or accept DF as well)
- Do not use Metropolis parameters, use mapping parameters
- allow for vector input in clusterelect
- fix AgeModel (StratMap) cluster logic
-->

<!-- The user may stumble across this vignette via a link or Google that places
it out of context.  Thus it's good practice to open with links to provide users
with necessary context; e.g. we should certainly link "StratoBayes" to the
main documentation home page.-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```
> StratoBayes is an R package for correlating and dating geological data.  
 - Installation instructions and documentation: [https://stratobayes.github.io/](https://stratobayes.github.io/).  
 - Overview of available vignettes: [Vignette index](https://stratobayes.github.io/vignettes/index).  

This guide walks you through the process of aligning geological data from multiple sites, using StratoBayes. Please [send us an email](mailto:contact@stratobayes.com?subject=StratoBayes%20R%20package) or raise an [issue on  Github](https://github.com/StratoBayes/StratoBayes/issues/new?title=Suggestion:+) if you encounter any problems with the software or have suggestions on improving this documentation. 

**Before proceeding, please ensure that you have [installed StratoBayes](https://stratobayes.github.io/software.html#installation)**. This vignette assumes that the package is installed and that you are running the provided code chunks in an R environment (using e.g., RStudio).

We start by loading the library:

```{r library}
library("StratoBayes")
```

Setting a random seed ensures the consistency of results:

```{r set-seed}
set.seed(0)
```

## Loading the data

The function `StratData()` is used for loading data containing the geological signal used for correlation, and, optionally, dates and partitioning information. The function can read CSV files directly, if the paths to the files are specified. Alternatively, R data.frames can be provided as input.

Here, we use the `signalData0` data.frame included in this package, containing a simulated signal recorded at two sites:

```{r signalData}
head(signalData0)
```

We can transform it into a `StratData` object, which tabulates the information in the format required to run a stratigraphic model, with `StratData()`:

```{r StratData}
stratData0 <- StratData(signal = signalData0)
```

Alternatively, we can read the data directly from a csv file:

```{r csv}
csvPath <- system.file("extdata", "signalData0.csv", package = "StratoBayes")
stratData0 <- StratData(signal = csvPath)
```

We can use the `plot()` method to visualize the raw data:

``` {r plot.StratData, fig.width=7, fig.height=3.5}
plot(stratData0)
```

## Model set-up

<!--Somewhere about here we should help a user to answer the question "Which models might be suitable for my own data, and how are they implemented in StratoBayes"-->


Now, we want to specify a stratigraphic correlation model that takes *Site&nbsp;1* as the reference section, and shifts and stretches the signal of *Site&nbsp;2* to achieve the best possible fit to the data of *Site&nbsp;1*.

The nature of the data informs the type of model we might use. In this case, we have no absolute age information, such as radiometric dates. We will thus fit the model on the "height" scale.  

> See [Age scale correlation and age modelling](https://stratobayes.github.io/vignettes/Tie-points.html) for an example with absolute ages.

Furthermore, we don't have information on potential sedimentation rate changes within the sections, e.g. due to lithology changes. We will therefore chose a simple model that assumes a uniform sedimentation rate within each section, but allows for sedimentation rates to differ from site to site.  

> See [Partitioning data](https://stratobayes.github.io/vignettes/Partitioning.html) for examples with sedimentation rate changes within sections.

The model used here will employ two parameters, denoting the offset of *Site&nbsp;2* relative to *Site&nbsp;1*, $\alpha_{site2}$, and the natural logarithm of the sedimentation rate of *Site&nbsp;2*, relative to *Site&nbsp;1*, $\ln(\gamma_{site2})$. The reference site *Site&nbsp;1* remains unchanged. In this model, a sedimentation rate of $\ln(\gamma_{site2}) = \ln(3) \approx 1.1$ would mean that the sedimentation rate of *Site&nbsp;2* is three times as high as that of *Site&nbsp;1*. Using the log-scale for the rate parameters ensures that the prior on the rate is symmetric; a doubling or halving of the rate has equivalent distances on the log-scale.


### Priors

As we are working in a Bayesian framework, we need to specify priors for those two parameters.
The helper function `StratModelTemplate()` will tell us which priors we need to specify, and provides the code needed to run the model.
When calling this function, we will specify that our model should be on the height scale (`alignmentscale = "height"`), since we don't have dates in our data set, and that our sedimentation rate model will assume a different sedimentation rate at each site (`sedModel = "site"`). We further specify that the offset of *Site&nbsp;2* relative to *Site&nbsp;1*, $\alpha_{site2}$, refers to the middle of *Site&nbsp;2* (`alphaPosition = "middle"`). 
Here, we will use a uniform prior on the $\alpha$ parameter, and a normal prior on the $\ln(\gamma)$ parameters:

``` {r PriorTemplate}
StratModelTemplate(
  stratData0,
  alignmentScale = "height",
  sedModel = "site",
  alphaPosition = "middle",
  alpha = "uniform",
  gammaLog = "normal"
)
```

Now, we can copy this template into our script and fill in the priors with appropriate values
Our reference section spans the interval from 0.25 to 18.7&nbsp;m. Using these as the `min` and `max` values for the uniform prior on $\alpha_{site2}$ means that the mid height of *Site&nbsp;2* may fall anywhere within the height range of *Site&nbsp;1*. 
For the normal prior on the log of the sedimentation rate of *Site&nbsp;1* relative to *Site&nbsp;2* we use a mean of 0 (a log sedimentation rate of 0 corresponds to a relative sedimentation rate of 1, i.e. no stretching or squeezing of the signal). We set the standard deviation to ln&nbsp;2, which places ~68% of the prior probability of this parameter on sedimentation rates between 0.5 and 2.0, with the remaining 32% being allocated to more extreme relative sedimentation rates.

``` {r priors}
stratPrior0 <- structure(list(
  "alpha_site2" = UniformPrior(min = 0.25, max = 18.7),
  "gammaLog_site2" = NormalPrior(mean = 0, sd = log(2))),
  class = c("StratPrior", "list"))
```

> **Default priors**  
> For quick exploratory analyses, automatically generated default priors can be used (instead of specifying own priors) by setting `priors = NULL` in the `StratModel()` function. See `?AutoPrior` for details.

### The model object 
Before being able to run the model, we have to create a model object. We can simply use the code suggested by the `StratModelTemplate` function, but because we have a small data set, we will decrease the number of knots for the splines to 10 (`nKnots = 10`), which will reduce the run time. Larger, more complicated data sets likely require more knots for the spline to fit the data well.

```{r model}
stratModel0 <- StratModel(stratData = stratData0,
                    priors = stratPrior0,
                    alignmentScale = "height",
                    sedModel = "site",
                    alphaPosition = "middle",
                    nKnots = 10)
```

## Running the model

We will estimate the parameter values with the `RunStratModel()` function by running a Markov chain Monte Carlo (MCMC) simulation to draw from the posterior of the model. `nIter` specifies the number of iterations to run. The more complicated the model is, the more iterations will be necessary to collect a sufficient number of independent samples from the posterior distribution. Here, we run it for 1,000 iterations. Samples from an MCMC are usually autocorrelated, so we use the `nThin` parameter to thin the samples and keep only every 5th sample by setting `nThin = 5`. `nRun` determines how many independent model runs will be conducted. To evaluate whether the model has converged, it is useful to have at least 3 independent model runs starting from different parameter values. Setting `runParallel` to `TRUE` allows to process each run on a separate CPU core, which can drastically improves overall run times. This requires installation of the `foreach` and the `doParallel` R packages. Note that progress bars don't work when doing parallel runs. Instead, StratoBayes attempts to display model alignments and iteration counts in a browser window when `runParallel = TRUE`. If you want to just do a single run, or several runs sequentially, set `runParallel` to `FALSE`. 

```{r, checkParallel}
runParallel <- requireNamespace("doParallel", quietly = TRUE) &&
  parallel::detectCores() > 1
runParallel
```

We run the model using the `RunStratModel()` function:

```{r, eval = TRUE, echo = FALSE}
result <- StratoBayes::stratPosterior0
```

```{r, eval = FALSE}
result <- RunStratModel(stratObject = stratData0,
                        stratModel = stratModel0,
                        nRun = 3,
                        nIter = 1000, 
                        nThin = 10,
                        runParallel = runParallel)
```

## Processing and analysing the results

`result` is now an object of class `StratPosterior`, which we can examine using the `print()`, `summary()`, and various `plot` methods. This particular result is also available for use in examples, without running the model, as package data `stratPosterior0`.

```{r print.StratPosterior}
result
```

### Checking for convergence

The first thing to do with the result of the model run is to check for convergence. Convergence means that the chains have reached a state were they are no longer influenced by their starting values, and they have thoroughly explored the parameter space. We also want to make sure that the chains are mixing well. This can be visually assessed with trace plots.

To plot the trace of $\alpha_{site2}$ and  $\ln \gamma_{site2}$, we can run:

```{r first-traceplot-code, eval = FALSE}
TracePlot(result, parameters = 1:2)
```

```{r plot-first-trace, fig.width = 5, fig.height = 4, echo = FALSE}
par(mar = c(4.1, 4.1, 1, 1))
TracePlot(result, parameters = 1:2)
```

Visual inspection of the trace plot shows that the chains are showing stable and overlapping patterns, suggesting that they have reached a stationary distribution and converged. For more information on assessing convergence, see [Gelman et al. 2016](http://www.stat.columbia.edu/~gelman/book/), Chapter 11.4.

### Plotting alignments

The `plot()` function displays the alignment corresponding to an approximate multivariate median of the converted heights (see [Eichenseer et al. 2025](https://egusphere.copernicus.org/preprints/2025/egusphere-2025-1355/)). If a Cluster analysis reveals more than one distinct possible alignment, the most likely alignment is shown by default.

```{r plot.StratPosterior, fig.width = 4, fig.height = 4.5}
plot(result, alignment = "all")
```

### Model run summary
The `summary()` function generates a summary of information from the model run. If more than one cluster of possible alignments has been identified, it will display information for each alignment cluster.  

```{r summary.StratPosterior}
summary(result)
```

### Stratigraphic mapping
With our age model, we can now convert any height from *Site&nbsp;2* to the height scale of our reference section *Site&nbsp;1* using the `StratMap()` function.
For example, to give us the heights in section_1 corresponding to a height of 1&nbsp;m in *Site&nbsp;2*, we can use:

```{r StratMap}
StratMap(result, heights = 1, site = "site2")
```

This tells us that a height of 1&nbsp;m at *Site&nbsp;2* would correspond to a mean height of `r signif(StratMap(result, heights = 1, site = "site2")$mean, 2)`&nbsp;m at *Site&nbsp;1*, for this alignment.

We can visualise the stratigraphic mapping of heights from *Site&nbsp;2* to the reference height scale from *Site&nbsp;1* using the `StratMapPlot()` function:

```{r stratmapplot-code, eval = FALSE}
StratMapPlot(result, site = "site2")
```

```{r plot-stratmapplot, fig.width = 3, fig.height = 2.9, echo = FALSE}
par(mar = c(4.1, 4.1, 1, 1))
StratMapPlot(result, site = "site2")
```


## Summary
<!-- [ Brief summary of what our model run has taught us and what we might conclude ]-->
Our model run identified the alignment that results in the best match of the proxy records from *Site&nbsp;1* and *Site&nbsp;1*, and estimated the uncertainty associated with that alignment. The median sedimentation rate of *Site&nbsp;2* relative to *Site&nbsp;1*, $\gamma_{site2} = \exp(\ln\gamma_{site2})$, has been estimated at `r round(exp(result$summary$alignment1$summary["gammaLog_site2", "50%"]), 1)`, meaning the analysis has "squeezed" the record of *Site&nbsp;2* by a factor of `r round(exp(result$summary$alignment1$summary["gammaLog_site2", "50%"]), 1)` to match that of *Site&nbsp;1*.

## Next steps
- [Partitioning data](https://stratobayes.github.io/vignettes/Partitioning.html) allows sedimentation rate changes and/or hiatuses, enabling more accurate correlations  
- [Tie points](https://stratobayes.github.io/vignettes/Tie-points.html) can improve correlations where available  
- [Age scale correlation and age modelling](https://stratobayes.github.io/vignettes/Tie-points.html) allows for dating sections using absolute age constraints
